
\documentclass [PhD] {uclathes}

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{linkcolor}{RGB}{0,0,102}
\usepackage{hyperref} 
\hypersetup{
	pdfstartview = {XYZ},
	colorlinks,
	allcolors = linkcolor,
	bookmarksopen,
	bookmarksnumbered
}
\usepackage{setspace}
\usepackage{silence}
\WarningFilter{caption}{Unknown} % captionpackage complains about thesis document class
\usepackage{caption}
\captionsetup{font={stretch=1}}
\usepackage{tabularx, colortbl}
\definecolor{mygray}{gray}{0.9}
\def \figscale {0.28}
\usepackage{multicol}
\usepackage{soul}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Usually things live in separate flies.
%
% \input {prelim}                           % preliminary page info

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                      %
%                          PRELIMINARY PAGES                           %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title          {Identifying genomic regulatory patterns \\
                 underlying complex phenotypes from heterogeneous data}
\author         {Brandon Jew}
\department     {Bioinformatics}
% Note:  degreeyear should be optional, but as of  5-Feb-96
% it seems required or you get a year of ``2''.   -johnh
\degreeyear     {2021}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chair          {Eran Halperin}
\chair          {Jae Hoon Sul}
\member         {Päivi Pajukanta}
\member         {Sriram Sankararaman}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\dedication     {\textsl{To my wife, \\
                my constant source of happiness and inspiration.}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\acknowledgments {First and foremost, I thank my advisors, Eran Halperin and Jae Hoon Sul. I am extremely grateful to have both as amazing mentors who provided me with exciting projects and opportunities to grow as a researcher. They both fostered welcoming and top-notch research groups that I am proud to be a part of. My \textsl{Halperin Lab} fleece jacket and \textsl{Sul Lab} mug will always be parts of my look as a scientist.   

I would also like to thank Eleazar Eskin, who roped me into bioinformatics in the first place and supported me through my PhD. Much of my work was done with Päivi Pajukanta, Noah Zaitlen, and Sriram Sankararaman. I am extremely grateful for their collaboration and mentorship. I thank Lana Martin and Robert Smith, who helped me immensely with funding and outreach projects during their time in Eleazar’s lab. I thank Sim-Lin Lau and Margaret Chu for their amazing support.

Many thanks to my lab mates in Eran's lab (Elior, Mike, Leah, Johnson, Liat, Ulzee, Brian, Nadav, Jeff) and Jae Hoon's Lab (Albert, Sarah, Lingyu). Also thanks to our shared lab space friends from Sriram's lab (Alec, Chris, Ruthie, Arun, Ariel, Ali). I really enjoyed talking too much during work hours and tracking down every opportunity for free food. 

Finally, I would like to thank my family for their unwavering support. I thank my wife, the amazing Dr. Maegan Lu. Over the past 8 years, she's been a constant source of support and inspiration to move forward. \\
%\\
%Chapter Two of this dissertation is a version of Brandon Jew, Marcus Alvarez, Elior Rahmani, Zong Miao, Arthur Ko, Kristina M. Garske, Jae Hoon Sul, Kirsi H. Pietiläinen, Päivi Pajukanta, Eran Halperin. Accurate estimation of cell composition in bulk expression through robust integration of single-cell information. \textsl{Nature Communications.} 2020.

%Chapter Three is a version of Brandon Jew, Jiajin Li, Sriram Sankararaman, Jae Hoon Sul. An efficient linear mixed model framework for meta-analytic association studies across multiple contexts. \textsl{To appear in WABI 2021}.

%Chapter Four is a version of Brandon Jew, Jae Hoon Sul, Noah Zaitlen. Selection contributes to skewed X chromosome inactivation across human tissues. \textsl{In preparation}.


This work was supported by the National Science Foundation Graduate Research Fellowship Program under Grant No. DGE-1650604.

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\vitaitem   {2017-2021}
                {PhD, Bioinformatics, University of California, Los Angeles, CA, USA}
\vitaitem   {2018-2021}
                {Graduate Research Fellow, National Science Foundation}
\vitaitem   {2013--2017}
                {BS, Biochemistry (minor in Bioinformatics), University of California, Los Angeles, CA, USA}
\vitaitem   {2014-2016}
                {Research Assistant, Loo Laboratory, University of California, Los Angeles, CA, USA}
\vitaitem   {2014}
                {Research Intern, Evans Laboratory, Salk Institute for Biological Studies}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
            
  \publication{* Denotes equal contribution}
  \publication{Aditya Gorla, \textbf{Brandon Jew}, Luke Zhang, Jae Hoon Sul. xGAP: A python based efficient, modular, extensible and fault tolerant genomic analysis pipeline for variant discovery. \textsl{Bioinformatics.} 2021.}
  \publication {David Goodman-Meza, Akos Rudas, Jeffrey N Chiang, Paul C Adamson, Joseph Ebinger, Nancy Sun, Patrick Botting, Jennifer A Fulcher, Faysal G Saab, Rachel Brook, Eleazar Eskin, Ulzee An, Misagh Kordi, \textbf{Brandon Jew}, Brunilda Balliu, Zeyuan Chen, Brian L Hill, Elior Rahmani, Eran Halperin, Vladimir Manuel. A machine learning algorithm to increase COVID-19 inpatient diagnostic capacity. \textsl{PLoS One.} 2020.}
  \publication {Zong Miao, Marcus Alvarez, Arthur Ko, Yash Bhagat, Elior Rahmani, \textbf{Brandon Jew}, Sini Heinonen, Karen L Mohlke, Markku Laakso, Kirsi H. Pietiläinen, Eran Halperin, Päivi Pajukanta. The causal effect of obesity on prediabetes and insulin resistance reveals the important role of adipose tissue in insulin resistance. \textsl{PLoS Genetics.} 2020.}
  \publication {Marcus Alvarez*, Elior Rahmani*, \textbf{Brandon Jew}, Kristina M Garske, Zong Miao, Jihane N Benhammou, Chun Jimmie Ye, Joseph R Pisegna, Kirsi H Pietiläinen, Eran Halperin, Päivi Pajukanta. Enhancing droplet-based single-nucleus RNA-seq resolution using the semi-supervised machine learning classifier DIEM. \textsl{Scientific Reports}. 2020.}
  \publication {\textbf{Brandon Jew}*, Marcus Alvarez*, Elior Rahmani, Zong Miao, Arthur Ko, Kristina M. Garske, Jae Hoon Sul, Kirsi H. Pietiläinen, Päivi Pajukanta, Eran Halperin. Accurate estimation of cell composition in bulk expression through robust integration of single-cell information. \textsl{Nature Communications.} 2020.}
    \presentation{\textbf{Brandon Jew}, Marcus Alvarez, Elior Rahmani, Zong Miao, Arthur Ko, Kristina M. Garske, Jae Hoon Sul, Kirsi H. Pietiläinen, Päivi Pajukanta, Eran Halperin. Accurate estimation of cell composition in bulk expression through robust integration of single-cell information. Platform presented at the 69th Annual Meeting of the American Society of Human Genetics, October 19, 2019, Houston, Texas.}
  \publication {Jennifer Zou, Farhad Hormozdiari, \textbf{Brandon Jew}, Stephane E Castel, Tuuli Lappalainen, Jason Ernst, Jae Hoon Sul, Eleazar Eskin. Leveraging allelic imbalance to refine fine-mapping for eQTL studies. \textsl{PLoS Genetics.} 2019.}
  \publication {Brian L Hill, Robert Brown, Eilon Gabel, Nadav Rakocz, Christine Lee, Maxime Cannesson, Pierre Baldi, Loes Olde Loohuis, Ruth Johnson, \textbf{Brandon Jew}, Uri Maoz, Aman Mahajan, Sriram Sankararaman, Ira Hofer, Eran Halperin. An automated machine learning-based model predicts postoperative mortality using readily-extractable preoperative electronic health record data. \textsl{British Journal of Anaesthesia.} 2019.}
  \publication {Jiajin Li, \textbf{Brandon Jew}, Lingyu Zhan, Sungoo Hwang, Giovanni Coppola, Nelson B Freimer, Jae Hoon Sul. ForestQC: quality control on genetic variants from next-generation sequencing data using random forest. \textsl{PLoS Computational Biology.} 2019.}
  \publication {\textbf{Brandon Jew}, Jae Hoon Sul. Variant calling and quality control of large-scale human genome sequencing data. \textsl{Emerging Topics in Life Sciences.} 2019.\\}
  
% \customCV{
% \begin{center}
%     SUBMITTED MANUSCRIPTS 
% \end{center}
%   \textbf{Brandon Jew}, Jae Hoon Sul, Noah Zaitlen. Selection contributes to skewed X chromosome inactivation across human tissues. \\
  
%   \noindent \textbf{Brandon Jew}, Jiajin Li, Sriram Sankararaman, Jae Hoon Sul. An efficient linear mixed model framework for meta-analytic association studies across multiple contexts. \\
  
%   \noindent Elior Rahmani*, \textbf{Brandon Jew}*, Eran Halperin. Differential DNA methylation at cell-type resolution: establishing rigorous guidelines.
% }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\abstract       {Large-scale transcriptomic datasets provide valuable opportunities to better understand the regulation of gene expression and its role in human health. However, these studies can be confounded by issues such as cell type heterogeneity. Furthermore, these datasets are growing extremely large with complex study designs, such as gene expression measured across a multitude of tissues, that must be accurately and efficiently modeled. Finally, better understanding of the mechanisms that influence gene regulation are required to integrate novel associations with biological understanding. In this dissertation, we introduce methods that address these issues in the analysis of tissue-level gene expression data. We present a method to accurately estimate cell type composition from these data by integrating single-cell information, as well as a scalable approach to model multi-tissue expression datasets and identify expression quantitative trait loci. We also present analyses of bulk expression data that validates a hypothesized mechanism of gene regulation that occurs in the general female population through non-random X chromosome inactivation. The work presented in this dissertation allows researchers to perform efficient and accurate analyses of gene expression data and provides additional insight into the mechanisms that underlie associations between genetics, transcriptomics, and complex phenotypes.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\begin {document}
\makeintropages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Ordinarily each chapter (at least) is in a separate file.
%
%\input {chapter1}                         % Chapter 1 of dissertation
%\input {chapter2}                         % Chapter 2
%\input {chapter3}                         % etc.
%\input {chapter4}
%\input {chapter5}
%\input {chapter6}
%\input {chapter7}
%\input {chapter8}

\input{chapter1/chapter1}

\input{chapter2/chapter2}

\input{chapter3/chapter3}

\input{chapter4/chapter4}
 \typeout{}
\bibliography{chapter1/chapter1, chapter2/chapter2, chapter3/chapter3, chapter4/chapter4}
\bibliographystyle{unsrt}

\end {document}

